Storm surge modelling in the Strait of Georgia

### 2014-07-24 RoughWork/manuscript.tex
Worked on past models of SoG in intro

\title{Storm surge modelling in the Strait of Georgia}
\author{Nancy Soontiens, Susan Allen, Doug Latornell, Kate Le Sou\"{e}f}
\date{\today}

@@

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are particularly susceptible to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly.
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here

@@

%overview oF SoG/Salish Sea models
Early efforts by \citet{crean1988numerical, stronach1993update} implemented a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. 
%tide models (Foreman), baroclinic (Masson, PNWL, Sutherland and MacCready)

>>>>

\title{Storm surge modelling in the Strait of Georgia}
\author{Nancy Soontiens, Susan Allen, Doug Latornell, Kate Le Sou\"{e}f, Idalia Machuca}
\date{\today}

@@

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are particularly susceptible to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here
%Need something about SoG size, geography, islands, sills, rivers....

@@

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988numerical, crean1988mathematical, stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinc version of their model examined residual currents and more. %elaborate!
 The baroclinic model of \citet{masson2004modelling} examined seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma vertical coordinate system leading to errors in diapycnal mixing due to steep topographics slopes. %fix this up still.
%tide models (Foreman), baroclinic (Masson, PNWL, Sutherland and MacCready)
